Stateless Model Checking with Data-Race Preemption Points

### 2015-11-19 pldi-16/design.tex
garth's comments for design/soundness

For kernel-level testing, we consider interrupt-disabling analogous to locking,
so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
%(to appropriate the names of the x86 instructions)
During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
Some kernels disable preemption without disabling interrupts,
which can be modelled the same way using manual annotations. %of that API.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
\quicksand~is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, we don't test every intermediate subset such as $\{lock,sti\}$,
which could potentially be improved in future work (\sect{\ref{sec:future}}).

@@

\begin{algorithm}[t]
	\SetKwInOut{Input}{Input}
	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
	\Input{$j_0$, the currently-running job}
	%\Input{$eta$, $j_0$'s predicted completion time}
	\Input{\PendingJobs, the list of pending jobs, sorted decreasingly by heuristic priority}
	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted increasingly by ETA}
	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
		// Common case: Job is expected to complete. \\
		return $j_0$
	}
	\ForEach{job $j_P \in$ \PendingJobs}{
		// Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
			return $j_P$
		}
	}
	%// no pending jobs; maybe resume a suspended job \\
	\ForEach{job $j_S \in$ \SuspendedJobs}{
		\If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
			$\land$
			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
			// If a subset of $j_S$ is also suspended, don't run the larger one first. \\
			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
				return $j_S$
			}
		}
	}
	return $j_0$ // ETA was bad, but no other job was better.
	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
	\label{alg:shouldworkblock}
\end{algorithm}

@@

During stateless MC, runtime data-race detection may find data-race candidates that we wish to investigate further.
Because data races indicate access pairs that can interleave at instruction granularity,
it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.

@@

With Iterative Deepening, this is a simple matter of creating a new state space with an additional PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
Rather than adding a single large state space, %configured to preempt on both involved instructions,
i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.

Furthermore, we do not always strictly increase the number of PPs when we find a new data race.
%When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
For each instruction involved in a data race, \quicksand~adds two new jobs:
a ``small'' job to preempt on that instruction only (line 5),
and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job (line 2).
%
Hence,
%together with the logic in \sect{\ref{sec:classifying}},
each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
%
The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
while the big job risks not completing in time,
the small job risks missing the data race entirely if the original PPs were required to expose it.
% TODO CAMREADY: Put numbers here.
In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
which motivates the need for Iterative Deepening to prioritize the jobs at runtime.

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
	\label{fig:new-dr-jobs}
\end{figure}

@@

\subsection{Heuristics}
% List of all heuristix:
% HOMESTRETCH - last 60sec of test, don't suspend
% ETA_THRESH - "to let its ETA stabilize"
% eta factor
% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
% priority change between suspected and confirmed dr
Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
to express how optimistic we are about the estimate's accuracy.
We use 2 as this scaling factor based on the results in \cite{estimation}.
%though we allow changing it via the command line.
We also include a heuristic to
%ignore ETAs entirely
never suspend jobs before they pass a certain threshold of interleavings tested,
for which we choose 32,
so that their ETAs have some time to stabilize.

We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
Jobs with both-order data-race PPs are prioritized higher,
because single-order candidates are more likely to be false-positives
(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
%so we prioritize jobs with both-order data-race PPs.
For single-order races, we do not add a PP for the later access at all;
should it be needed, preempting on the first access will suffice to upgrade the race to both-order.

>>>>

For kernel-level testing, we consider interrupt-disabling analogous to locking,
so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
%(to appropriate the names of the x86 instructions)
During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
Some kernels disable preemption without disabling interrupts,
which can be communicated to the MC using manual annotations. %of that API.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
\quicksand~is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, we don't test every intermediate subset such as $\{lock,sti\}$,
which could potentially be improved in future work (\sect{\ref{sec:future}}).

@@

\begin{algorithm}[t]
	\SetKwInOut{Input}{Input}
	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
	\Input{$j_0$, the currently-running job}
	%\Input{$eta$, $j_0$'s predicted completion time}
	\Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ TimeLeft()}{
		// Common case: Job is expected to complete. \\
		return $j_0$
	}
	\ForEach{job $j_P \in$ \PendingJobs}{
		// Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
			return $j_P$
		}
	}
	%// no pending jobs; maybe resume a suspended job \\
	\ForEach{job $j_S \in$ \SuspendedJobs}{
		\If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
			$\land$
			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
			// If a subset of $j_S$ is also suspended, don't run the larger one first. \\
			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
				return $j_S$
			}
		}
	}
	return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
	\caption{Suspending exploration of a state space in favour of a potentially smaller one.}
	\label{alg:shouldworkblock}
\end{algorithm}

@@

During stateless MC, runtime data-race detection may find data-race candidates that we wish to investigate further.
Because data races indicate access pairs that can interleave at instruction granularity,
it is logical to re-execute the test and issue preemptions just before those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.

@@

With Iterative Deepening, this is a simple matter of creating a new state space with an additional PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
We call these {\em data-race PPs}.
Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
Rather than adding a single large state space, %configured to preempt on both involved instructions,
i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.

Furthermore, we do not always strictly increase the number of PPs when we find a new data race.
%When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
For each instruction involved in a data race, \quicksand~adds two new jobs:
a ``small'' job to preempt on that instruction only (line 5),
and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job (line 2).
%
Hence,
%together with the logic in \sect{\ref{sec:classifying}},
each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
%
The rationale of spawning multiple jobs is that we don't know in advance which will be most fruitful: %which will be most fruitful is not known in advance:
while the big job risks not completing in time,
the small job risks missing the data race entirely if the original PPs were required to expose it.
% TODO CAMREADY: Put numbers here.
In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs.
%which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
This phenomenon motivates Iterative Deepening to prioritize the jobs at run-time.

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
		When an access pair is reported as a data race candidate, we generate a new PP for each access, and add new jobs corresponding to different combinations of those with the existing PPs.}
	\label{fig:new-dr-jobs}
\end{figure}

@@

\subsection{Heuristics}
% List of all heuristix:
% HOMESTRETCH - last 60sec of test, don't suspend
% ETA_THRESH - "to let its ETA stabilize"
% eta factor
% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
% priority change between suspected and confirmed dr
Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
to express how pessimistic we are about the estimate's accuracy.
We use a scaling factor defaulting to 2 based on the results in \cite{estimation}.
%though we allow changing it via the command line.
We also include a heuristic to
%ignore ETAs entirely
never suspend jobs before they pass a certain threshold of interleavings tested,
for which we choose 32,
so that their ETAs have some time to stabilize.

We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
Single-order candidates are more likely to be false positives,
although preempting during the access itself is necessary to say for sure.
Hence, we add PPs for both types of candiates, and heuristically prioritize jobs with both-order data-race PPs
(indicated by the HeuristicPriority($\alpha$) call in Algorithm~\ref{alg:handledatarace}).
%Jobs with both-order data-race PPs are prioritized higher,
%because single-order candidates are more likely to be false-positives.
%(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
%so we prioritize jobs with both-order data-race PPs.
For single-order races, we do not initially add a PP for the later access at all:
if preempting on the first access can reorder the race, it will be upgraded to both-order in the new state space, and we will add the second PP then.
%should it be needed, preempting on the first access will suffice to upgrade the race to both-order.

### 2015-11-19 pldi-16/implementation.tex
garth's comments for design/soundness

We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
\landslide~implements DPOR \cite{dpor},
%\cite{dpor},
state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
% this line can be cut if space is needed
It can test both user-level and kernel-level code, although is limited to timer nondeterminism.
% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
Its bug-detection metrics are assertion failure, deadlock, segfault, heap checking \cite{valgrind}, and a heuristic infinite loop check.

{\bf Restricting PPs with stack trace predicates.}
Most MC tools preempt indiscriminately on any sync API call, regardless of the call-site.
However, when testing a particular module in a large codebase,
the user is likely uninterested in PPs arising from other modules.
\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
%The {\tt without\_function} directive works similarly, but as a blacklist.
The {\tt without\_function} directive is the dual of {\tt within\_function}, indicating a blacklist.
Multiple invocations can be used; later ones take precedence.
%\cite{landslide} provides further detail on this feature.

{\bf Data races in lock implementations.}
Data race tools in prior work \cite{tsan,portend} recognize the implementations of sync primitives to avoid spuriously flagging memory accesses therein. %resulting from the lock implementation itself.
The assumption that the locks are already correct enables productive data-race analysis on the rest of the codebase.
Otherwise, with testing limited to one execution,
%even if one wishes to test for lock bugs,
data-race analysis would flag every access pair in the lock implementation. %, requiring human attention to verify.
However, Iterative Deepening removes the need for human attention to verify them. %can automatically verify a large quantity of data-race candidates as benign.
Hence, we extended \landslide~with an option to make its data race analysis consider accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()},
which we use in \sect{\ref{sec:testsuite}}.
%(Accesses from other sync functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)

@@

\quicksand~is an independent program that wraps the execution of several \landslide~instances.
The implementation is roughly 3000 lines of C.
%It uses a thread pool to schedule the available state spaces,
%sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
%Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
The interface to \landslide~ has two parts. %, which any similar MC could implement, has two parts.
First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
% can lose this line due to space
among other options such as mutex-testing mode,
passed as an argument to \landslide.
Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing. %on a FIFO pipe.
\landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
and whenever it finds a new data-race candidate or bug.
\quicksand~in turn replies whether to suspend/resume due to too high ETA, or quit due to timeout.
We suspend jobs simply by making \landslide~wait on a message-passing call.
Should \quicksand~later resume a suspended job, we send a message to continue,
resuming the \landslide~instance where it left off.
%otherwise, we resume it only after time runs out, causing it to exit.

>>>>

We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
\landslide~implements DPOR \cite{dpor},
%\cite{dpor},
state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
% this line can be cut if space is needed
It can test both user-level and kernel-level code (although it is limited to timer nondeterminism).
% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
Its bug-detection metrics are assertion failure, deadlock, segfault, heap checking \cite{valgrind}, and a heuristic infinite loop check.

{\bf Restricting PPs with stack trace predicates.}
When testing a particular module in a large codebase,
the user is likely to be uninterested in PPs arising from other modules.
Rather than preempting indiscriminately on any synchronization call, regardless of the call-site,
\landslide~provides a configuration command, {\tt within\_function}, for identifying which call-sites matter.
%Most MC tools preempt indiscriminately on any synchronization call, regardless of the call-site.
%However, when testing a particular module in a large codebase,
%the user is likely to be uninterested in PPs arising from other modules.
%\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
%Similarly, the {\tt without\_function} directive indicates a blacklist,
%serving as the dual of {\tt within\_function}.
The {\tt without\_\allowbreak{}function} directive is the dual of {\tt within\_function}, indicating a blacklist.
We use this feature to restrict the scope of some tests in our evaluation (\sect{\ref{sec:testsuite}}).
%Multiple invocations can be used; later ones take precedence.
%\cite{landslide} provides further detail on this feature.

{\bf Data races in lock implementations.}
Data race tools in prior work \cite{tsan,portend} recognize the implementations of synchronization primitives to avoid spuriously flagging memory accesses therein. %resulting from the lock implementation itself.
Assuming that the lock implementation is already correct enables more productive data-race analysis on the rest of the codebase.
Otherwise, with testing limited to one execution,
%even if one wishes to test for lock bugs,
data-race analysis would flag every access pair in the lock implementation. %, requiring human attention to verify.
However, Iterative Deepening removes the need for human attention to verify them. %can automatically verify a large quantity of data-race candidates as benign.
Hence, we extended \landslide~to
%with an option
optionally make its data race analysis consider accesses from {\tt mutex\_lock()} and {\tt mutex\_unlock()},
which we use in \sect{\ref{sec:testsuite}}.
%(Accesses from other synchronization functions, such as {\tt cond\_wait()}, would either be included already, or be protected by an internal mutex.)

@@

\quicksand~is an independent program that wraps the execution of several \landslide~instances.
The implementation is roughly 3000 lines of C.
%It uses a thread pool to schedule the available state spaces,
%sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
%Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
The interface with the MC has two parts. %, which any similar MC could implement, has two parts.
First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
% can lose this line due to space
plus other MC-specific options such as our modifications to \landslide~for testing mutexes. %mutex-testing mode,
%passed as an argument to \landslide.
Then, a dedicated \quicksand~thread communicates with the MC process via message-passing. %on a FIFO pipe.
%\landslide~messages \quicksand~
The MC messages after testing each new interleaving to report updated progress and ETA
and whenever a new data-race candidate or bug is found.
\quicksand~in turn replies whether to suspend/resume due to too high ETA, or quit due to timeout.
We suspend jobs simply by making the MC wait on a message-passing call.
Should \quicksand~later re-schedule a suspended job, we send a message to continue,
resuming it where it left off.
%otherwise, we resume it only after time runs out, causing it to exit.

### 2015-11-19 pldi-16/soundness.tex
garth's comments for design/soundness

\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
\begin{theorem}
If a bug can be exposed by any thread interleaving possible by preempting on any instruction,
Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
\end{theorem}
\begin{proof}
The proof has two parts:
first, we show that preempting on data-racing instructions and synchronization API boundaries suffices to test all possible program behavior;
second, we show that Iterative Deepening will eventually detect all such data races.
Given a PP $p$, let $\ppnext{p}$ denote the next transition after $p$ executed by the thread which ran immediately before $p$,
let $\ppinstr{p}$ denote the first instruction of $\ppnext{p}$,
and let $\ppothers{p}$ denote the transitions by other threads between $p$ and $\ppnext{p}$.

\begin{lemma}[Equivalence of non-data-race PPs]
For any thread interleaving possible by preempting on any instruction,
there exists an equivalent interleaving which uses only data-race PPs and sync API PPs.
	\label{lem:relevant}
\end{lemma}

Let $p$ be the first PP in the given interleaving such that $\ppinstr{p}$ is not a data race with $\ppothers{p}$ nor is a sync API boundary.
Because $\ppinstr{p}$ is not a sync API boundary,
no lock can be held during $\ppothers{p}$ that was also held by the first thread across $p$.
Hence, because $\ppinstr{p}$ is not a data race, it cannot be a shared memory conflict with $\ppothers{p}$ at all.
Let $i$ be the first instruction among $\ppnext{p}$ which is such a conflict, or a sync API boundary.
If $i$ is a shared memory conflict, it must be a data race, for the same reasoning as above.
We modify the input interleaving by reordering $\ppinstr{p}$ until $i$, not including $i$, to before $\ppothers{p}$.
By the soundness of DPOR \cite{dpor}, this is equivalent to the input interleaving.
In other words, we have transformed $p$ into $p'$ such that $\ppnext{p'} = i$, which is a data race or sync API boundary.
All PPs in the input trace can be inductively converted in the same manner.

\begin{definition}[Reachable data race]
A data race candidate (or associated PP) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
\end{definition}
Initially, the statically-available sync API PPs are reachable. Reachability of data-race PPs is transitive.
\begin{lemma}[Saturation of data-race PPs]
	Given any interleaving comprising only data-race PPs and sync API PPs, all such PPs are reachable.
	\label{lem:saturation}
\end{lemma}

@@

To conclude,
for any possible interleaving, Lemma \ref{lem:relevant} provides an equivalent one with only data-race and sync API PPs,
and Lemma \ref{lem:saturation} proves all involved PPs are reachable.
Hence, Iterative Deepening will eventually test a state space containing the equivalent buggy interleaving.
\end{proof}

@@

We identify a particular class of false positive data-race candidate in which the memory is recycled by {\tt malloc} between the two accesses.
Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a data race.
We term this a {\em malloc-recycle data race}.
To the human eye, this is obviously a false positive: reordering lines 4-7 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
Here, Thread 2's logic usually corresponds to the initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.

@@

It is quite simple to mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address.
We implemented this check by adding a generation counter to \landslide's heap tracking:
each allocation is given a unique ID,
and when evaluating whether two heap accesses can race,
the IDs of their containing blocks must match.
However, when limited to a single execution, suppressing any data race matching this pattern is unsound.
Consider the more unusual program in Figure~\ref{fig:recycle-bug}:
Now, the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
Hence, reordering lines 6-7 before line 3 will cause {\tt x} and {\tt x2} to race.

@@

\begin{proof}
By definition of the malloc-recycle pattern,
any such program must contain an access {\tt x1} by one thread T1,
followed by a {\tt free} and a {\tt malloc} possibly by either thread,
followed by an access {\tt x2} by the other thread T2. % not depending on the result of the middle malloc.
For brevity we say that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}; the other cases are similar.
We also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
hence, there must also be some ``publish'' action {\tt p} by T1 which communicates the address to T2.
Because this is a true data race, {\tt p} must occur before {\tt x1}, as {\tt x2} cannot be reordered before {\tt p}.

We now show that a PP will be identified during T1 between {\tt p} and {\tt x1}.
The publish action must involve some thread communication, whether through a shared data structure or message-passing API.
If locking or message-passing is used, our set of hard-coded PPs suffices.
Otherwise, {\tt p} (and the corresponding read by T2) will be a data race, although it may itself be a malloc-recycle race.
In this case we use induction on the pointer chain leading to {\tt x}:
in the base case, {\tt p} is global or obtained via message-passing,
and in the inductive step, DPOR will reorder threads sufficiently to identify the PP on {\tt p}.
Hence there will be a PP between {\tt p} and {\tt x1} no matter the mode of communication.

With this PP, DPOR will reorder {\tt x2} before {\tt x1}, while not changing {\tt x2}'s location.
As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
Hence {\tt x1} and {\tt x2} will be in the same allocation;
hence the accesses can race without fitting the malloc-recycle pattern.
% Mario-man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
% This spells QED so we are done.
\end{proof}
\renewcommand\proofname{Proof}

Note that we do not require PPs on the internal locking of {\tt malloc} or {\tt free},
which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}) to reduce state space size.

>>>>

\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
\begin{theorem}
If a bug can be exposed by any thread interleaving possible by preempting on all instructions,
Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
\end{theorem}
\begin{proof}
The proof has two parts:
first, we show that preempting on data-racing instructions and synchronization API boundaries suffices to test all possible program behavior;
second, we show that Iterative Deepening will eventually detect all such data races.
Given a PP $p$, let $\ppnext{p}$ denote the next transition after $p$ executed by the thread which ran immediately before $p$,
let $\ppinstr{p}$ denote the first instruction of $\ppnext{p}$,
and let $\ppothers{p}$ denote the transitions by other threads between $p$ and $\ppnext{p}$.

\begin{lemma}[Equivalence of non-data-race PPs]
For any thread interleaving possible by preempting on any instruction,
there exists an equivalent interleaving which uses only data-race PPs and synchronization API PPs.
	\label{lem:relevant}
\end{lemma}

Let $p$ be the first PP in the given interleaving such that $\ppinstr{p}$ is not a data race with $\ppothers{p}$ nor is a synchronization API boundary.
Because $\ppinstr{p}$ is not a synchronization boundary,
no lock can be held during $\ppothers{p}$ that was also held by the first thread across $p$.
Hence, because $\ppinstr{p}$ is not a data race, it cannot be a shared memory conflict with $\ppothers{p}$ at all.
Let $i$ be the first instruction among $\ppnext{p}$ which is such a conflict, or a synchronization boundary.
If $i$ is a shared memory conflict, it must be a data race, for the same reasoning as above.
We modify the input interleaving by reordering $\ppinstr{p}$ until $i$, not including $i$, to before $\ppothers{p}$.
By the soundness of DPOR \cite{dpor}, this is equivalent to the input interleaving.
In other words, we have transformed $p$ into $p'$ such that $\ppnext{p'} = i$, which is a data race or synchronization boundary.
All PPs in the input trace can be inductively converted in the same manner.

\begin{definition}[Reachable data race]
A data race candidate (and associated PPs) is reachable if it will be identified by an MC configured to preempt only on already-reachable PPs.
\end{definition}
Initially, the statically-available synchronization API PPs are reachable. Reachability of data-race PPs is transitive.
\begin{lemma}[Saturation of data-race PPs]
	Given any interleaving comprising only data-race PPs and synchronization API PPs, all involved PPs are reachable.
	\label{lem:saturation}
\end{lemma}

@@

To conclude,
for any possible interleaving, Lemma \ref{lem:relevant} provides an equivalent one with only data-race and synchronization PPs,
and Lemma \ref{lem:saturation} proves all involved PPs are reachable.
Hence, Iterative Deepening will eventually test a state space containing the equivalent buggy interleaving.
\end{proof}

@@

We identify a particular class of false positive data-race candidates in which the memory was recycled by {\tt malloc} between the two accesses.
Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behavior.
If the {\tt malloc} on line 4 returns the same address passed to {\tt free} on line 2, then lines 1 and 7 will be flagged as a data race.
We call this a {\em malloc-recycle data race}.
To the human eye, this is obviously a false positive: reordering lines 4-7 before lines 1-2 will change {\tt malloc}'s return value, causing {\tt x} and {\tt y} to no longer collide.
Here, Thread 2's logic usually corresponds to an initialization pattern \cite{eraser}, but for generality we have added a {\tt publish} action on line 6.

@@

We mechanically recognize when {\tt x} and {\tt y} correspond to different abstract allocations despite colliding on address,
%We implemented this check
by adding a generation counter to \landslide's heap tracking:
each allocation is given a unique ID,
and when evaluating whether two heap accesses can race,
the IDs of their containing blocks must match.
%However,
When limited to a single test execution, suppressing any data race matching this pattern is unsound.
Consider the more unusual program in Figure~\ref{fig:recycle-bug},
in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
Here, reordering lines 6-7 before line 3 will allow {\tt x} and {\tt x2} to race.

@@

\begin{proof}
By definition of the malloc-recycle pattern,
any such program must contain an access $a_1$ by one thread T1,
followed by a {\tt free} and a {\tt malloc} possibly by either thread,
followed by an access $a_2$ by the other thread T2. % not depending on the result of the middle malloc.
Without loss of generality, we say that T1 performs the {\tt free} and T2 the subsequent {\tt malloc}. %; the other cases are similar.
We also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
hence, there must also be some ``publish'' action $p$ by T1 which communicates the address to T2.
Because this is a true data race, $p$ must occur before $a_1$, as $a_2$ cannot be reordered before $p$.

We now show that a PP will be identified during T1 between $p$ and $a_1$.
The publish action must involve some thread communication, whether through a shared data structure or message-passing API.
If locking or message-passing is used, our set of hard-coded PPs suffices to provide a PP.
Otherwise, $p$ (and the corresponding read by T2) will be a data race, although that may itself be a malloc-recycle race.
In this case we use induction on the pointer chain leading to the shared address containing $p$:
in the base case, $p$ is communicated via global data or message-passing,
and in the inductive step, DPOR will reorder threads sufficiently to identify the PP on $p$.
Hence there will be a PP between $p$ and $a_1$ no matter the mode of communication.

With this PP, DPOR will reorder $a_2$ before $a_1$, while not changing $a_2$'s location.
As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
Hence $a_1$ and $a_2$ will be in the same allocation;
hence the accesses can race without fitting the malloc-recycle pattern.
% Mario-man is very very hunger from not having enough plumming jobs, so his Quest for Eat and Dollars.
% This spells QED so we are done.
\end{proof}
\renewcommand\proofname{Proof}

Note that we do not require PPs on the internal locking of {\tt malloc} or {\tt free},
which are ideal candidates to ignore via {\tt without\_function} (\sect{\ref{sec:landslide}}, \sect{\ref{sec:future}}) to reduce state space size.
